A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-03-01 sections/Metatheory.tex
S4 completeness corollary; statistics table

As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.
\jimmy{Cite DK's proof as the sketch.}

@@

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgment chains:
\begin{enumerate}
    \item If $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$ and $\|\Om\| \vdash A' \le B'$
        then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$ and $\|\Om\| \vdash e \Lto A'$
        then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Gm \sto \Om$ and exists $A$ s.t. $\|\Om\| \vdash e \To A$
        then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Gm \Vdash \appInfAlg{A}{e}[a][1 \to 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \to 1]$
        and exists $C$ s.t. $\|\Om\| \vdash \appInf{A'}{e}{C}$
        \\
        then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}

@@

% ---------------------------------------------
\subsection{Abella and Proof Statistics}
We have chosen the Abella (v2.0.7-dev) proof assistant~\cite{AbellaDesc} to develop our formalization.
Abella is designed to help such formalizations for programming languages,
due to its built-in support for variable bindings and
the $\lambda$-tree syntax~\cite{miller2000abstract}, i.e. a form of HOAS.
Nominal variables, or $\nabla$-quantified variables, are used as an unlimited name supply,
which supports explicit freshness control and substitutions.

@@

\begin{figure}[ht]
\centering\begin{tabular}{|c|c|c|l|}\hline
    File(s) & SLOC & \#Theorems & Description\\\hline
    olist.thm, nat.thm  &  309 & 56  & Basic data structures\\\hline
    typing.thm          &  238 & 5   & Declarative \& algorithmic system, debug examples\\\hline
    decl.thm            &  226 & 33  & Basic declarative properties\\\hline
    order.thm           &  235 & 27  & The $|\cdot|_\forall$ measure; decl. subtyping strenthening\\\hline
    alg.thm             &  655 & 72  & Basic algorithmic properties\\\hline
    trans.thm           &  628 & 53  & \makecell[l]{Transfer and its basic properties \\
                            Lemmas~\ref{lem:insert}, \ref{lem:extract}}\\\hline
    declTyping.thm      &  944 & 72  & \makecell[l]{Additional properties for declarative typing \\
                            Lemmas~\ref{lem:inv_allR}, \ref{lem:inv_chkAll}, \ref{lem:inv_chkLam}}\\\hline
    soundness.thm       & 1069 & 70  & Soundness theorem; aux. lemmas on transfer\\\hline
    depth.thm           &  206 & 14  & \makecell[l]{The $|\cdot|_\to$ measure; \\
                            Lemma~\ref{lem:prune_inst}}\\\hline
    \makecell{dcl.thm, \\completeness.thm} & 1458 & 73  &
                            Completeness theorem; aux. lemmas and relations\\\hline
    inst\_decidable.thm &  830 & 42  & Lemma~\ref{lemma:inst:decidable}\\\hline
    decidability.thm    &  885 & 50  & Decidability theorem\\\hline
    Total               & 7683 & 567 & (51 definitions in total)\\\hline
\end{tabular}
\caption{Statistics for the proof scripts}
\label{fig:proof_statistics}
\end{figure}

>>>>

As the algorithmic system introduces existential variables,
a declarative rule may correspond to multiple algorithmic rules,
and thus we analyse each of the possible cases.
% \jimmy{Cite DK's proof as the sketch.}
% I later think it might not be needed; the induction is not that hard to think of

@@

The following corollary is derived immediately from Theorem~\ref{thm:completeness}.
\begin{corollary}[Completeness, single judgment form]
Given \emph{wf }$\Gm$ containing no judgment chains:
\begin{enumerate}
    \item If $\Om \vdash A' \le B'$ and $\Gm \Vdash A \le B \sto \Om \Vdash A' \le B'$
        then $\Gm \Vdash A \le B \redto \nil$;
    \item If $\Om \vdash e \Lto A'$ and $\Gm \Vdash e \Lto A \sto \Om \Vdash e \Lto A'$
        then $\Gm \Vdash e \Lto A \redto \nil$;
    \item If $\Om \vdash e \To A$ and $\Gm \sto \Om$
        then $\Gm \Vdash e \To_a 1 \le 1 \redto \nil$;
    \item If $\Om \vdash \appInf{A'}{e}{C}$ and
        $\Gm \Vdash \appInfAlg{A}{e}[a][1 \to 1] \sto \Om \Vdash \appInfAlg{A'}{e}[a][1 \to 1]$
        \\
        then $\Gm \Vdash \appInfAlg{A}{e}[a][1 \le 1] \redto \nil$.
\end{enumerate}
\end{corollary}

@@

% ---------------------------------------------
\subsection{Abella and Proof Statistics}
We have chosen the Abella (v2.0.7-dev
\footnote{A develop version is used because the dev just fixed a serious bug that
accepts a simple proof of false, which also affects the completeness of our proof.
The author compiled the scripts against commit 92829a of the GitHub repository of Abella.}
) proof assistant~\cite{AbellaDesc} to develop our formalization.
Abella is designed to help such formalizations for programming languages,
due to its built-in support for variable bindings and
the $\lambda$-tree syntax~\cite{miller2000abstract}, i.e. a form of HOAS.
Nominal variables, or $\nabla$-quantified variables, are used as an unlimited name supply,
which supports explicit freshness control and substitutions.

@@

\begin{figure}[ht]
\centering\begin{tabular}{|c|c|c|l|}\hline
    File(s) & SLOC & \#Theorems & Description\\\hline
    olist.thm, nat.thm  &  309 & 56  & Basic data structures\\\hline
    typing.thm          &  238 & 5   & Declarative \& algorithmic system, debug examples\\\hline
    decl.thm            &  226 & 33  & Basic declarative properties\\\hline
    order.thm           &  235 & 27  & The $|\cdot|_\forall$ measure; decl. subtyping strenthening\\\hline
    alg.thm             &  655 & 72  & Basic algorithmic properties\\\hline
    trans.thm           &  628 & 53  & \makecell[l]{Transfer and its basic properties \\
                            Lemmas~\ref{lem:insert}, \ref{lem:extract}}\\\hline
    declTyping.thm      &  944 & 72  & \makecell[l]{Non-overlapping declarative system \\
                            Lemmas~\ref{lem:inv_allR}, \ref{lem:inv_chkAll}, \ref{lem:inv_chkLam}}\\\hline
    soundness.thm       & 1069 & 70  & Soundness theorem; aux. lemmas on transfer\\\hline
    depth.thm           &  206 & 14  & The $|\cdot|_\to$ measure; Lemma~\ref{lem:prune_inst}\\\hline
    dcl.thm             &  383 & 12  & Non-overlapping declarative worklist \\\hline
    completeness.thm    & 1075 & 61  &
                            Completeness theorem; aux. lemmas and relations\\\hline
    inst\_decidable.thm &  830 & 42  & Other measures; Lemma~\ref{lemma:inst:decidable}\\\hline
    decidability.thm    &  885 & 50  & Decidability theorem\\\hline
    Total               & 7683 & 567 & (51 definitions in total)\\\hline
\end{tabular}
\caption{Statistics for the proof scripts}
\label{fig:proof_statistics}
\end{figure}
